Hypothesis tests for multiple responses regression: effect of probiotics on addiction and binge eating disorder.

### 2023-02-14 5_artigos/statistical-methods-in-medical-research/submission/SOURCE/paper-smmr-lineu.tex
atualiza commit

\documentclass[AMA,STIX1COL]{WileyNJD-v2}

\articletype{RESEARCH ARTICLE}%

\received{Day April 2016}
\revised{6 June 2016}
\accepted{6 June 2016}

\raggedbottom

@@

@@

\author[1]{Lineu Alberto Cavazani de Freitas}

\author[2]{Ligia de Oliveira Carlos}

\author[2]{Antônio Carlos Ligocki Campos}

\author[3]{Wagner Hugo Bonat}

\authormark{FREITAS \textsc{et al}}

\address[1]{\orgdiv{Department of Informatics}, \orgname{Paraná Federal University}, \orgaddress{\state{Paraná}, \country{Brazil}}}

\address[2]{\orgdiv{Department of Surgical Clinic}, \orgname{Paraná Federal University}, \orgaddress{\state{Paraná}, \country{Brazil}}}

\address[3]{\orgdiv{Department of Statistics}, \orgname{Paraná Federal University}, \orgaddress{\state{Paraná}, \country{Brazil}}}

\corres{Lineu Alberto Cavazani de Freitas, Department of Informatics, Paraná Federal University, R. Evaristo F. Ferreira da Costa, 383-391, Jardim das Américas, Curitiba, Paraná, 82590-300, Brazil. \email{lineuacf@gmail.com}}

\abstract[Summary]{

@@

}

@@

% \jnlcitation{\cname{%
% \author{Freitas LAC}, 
% \author{Carlos LO},
% \author{Campos ACL} and 
% \author{Bonat WH}} (\cyear{2022}), 
% \ctitle{Wald test to evaluate regression and dispersion parameters in multivariate covariance generalized linear models}, \cjournal{Statistics in Medicine}}

@@

%\footnotetext{\textbf{Abbreviations:} ANA, anti-nuclear antibodies; APC, antigen-presenting cells; IRF, interferon regulatory factor}

@@

According to Meinert and Tonascia \cite{meinert1986clinical} "a clinical trial is a planned experiment designed to assess the efficacy of a treatment in man by comparing the outcomes in a group of patients treated with the test with those observed in a comparable group of patients receiving a control treatment where patients in both groups are enrolled, treated, and followed over the same time period".

@@

Similarly to the aforementioned articles, we are interested in the analysis of multiple outcomes in the context of longitudinal data analysis. The study we shall describe was conducted with the objective of evaluating the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. This is a clinical trial with two response variables: a score that characterizes compulsion and the number of symptoms presented that characterize addiction. Both responses are limited counts and for analysis purposes have been transformed to unit range.

@@

@@

We present the dataset in \autoref{sec2}. In \autoref{sec3} we present the review of the general structure and estimation of the parameters of a McGLM, based on the ideas of Bonat and Jørgensen \cite{Bonat16}. In \autoref{sec4} our proposal is presented with the details of the Wald test to evaluate assumptions about parameters of a McGLM. In \autoref{sec5} we present the results of the performance evaluation of the proposed test based on a simulation study. In the \autoref{sec6} we apply the model and present the main results of the analysis of the data from the clinical study that aims to evaluate the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. Finally, the main results are discussed in \autoref{sec7}, including some directions for future investigations. The data sets and R codes are available online and can be assessed in \href{https://github.com/lineu96/paper-htmcglm}{https://github.com/lineu96/paper-htmcglm}.

@@

\begin{figure}[h]
\centerline{\includegraphics[scale = 0.9]{FIGURE1}}
\caption{Exploratory graphic analysis: (a) YFAS histogram, (b) YFAS boxplots as a function of group, (c) YFAS boxplots as a function of moment, (d) BES histogram, (b) BES boxplots as a function of group, (c) BES boxplots as a function of moment. The asterisk in the boxplots indicates the mean.\label{fig1}}
\end{figure}

@@

\noindent where the $g_r()$ are standard link functions; $\boldsymbol{X}_r$ denotes a $N \times k_r$ design matrix; $\boldsymbol{\beta}_r$ denotes a  $k_r \times 1$ vector of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{ \Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\boldsymbol{\Sigma}}_R^\top)$ denotes the generalized Kronecker product \cite{martinez13}, the matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. The operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is an identity matrix $N \times N$.

@@

McGLMs' second-moment assumptions allow us to spplit the parameter vector into two sets: $\boldsymbol{\theta} = (\boldsymbol{\beta}^{\top}, \boldsymbol{\lambda}^{\top})^ {\top}$. So $\boldsymbol{\beta} = (\boldsymbol{\beta}_1^\top, \ldots, \boldsymbol{\beta}_R^\top)^\top$ is a $K \times 1$ vector of regression parameters and $\boldsymbol{\lambda} = (\rho_1, \ldots, \rho_{R(R-1)/2}, p_1, \ldots, p_R, \boldsymbol{\tau}_1^\top , \ldots, \boldsymbol{\tau}_R^\top)^\top$ is a $Q \times 1$ vector of dispersion parameters. Also, $\mathcal{Y} = (\boldsymbol{Y}_1^\top, \ldots, \boldsymbol{Y}_R^\top)^\top$ denotes the stacked vector of order $NR \times 1$ from the matrix of response variables $\boldsymbol{Y}_{N \times R}$ and $\mathcal{M} = (\boldsymbol{\mu}_1^\top, \ldots, \boldsymbol{\mu}_R^ \top)^\top$ denotes the stacked vector of order $NR \times 1$ of the expected value matrix $\boldsymbol{M}_{N \times R}$.

@@

\noindent In this way, the generalization of the Wald test statistic to verify the validity of a hypothesis about parameters of a McGLM is given by:

@@

\noindent where $W \sim \chi^2_s$, that is, regardless of the number of parameters in the hypotheses, the test statistic $W$ is a single value that asymptotically follows the $\chi^2$ distribution with degrees of freedom given by the number of constraints, that is, the number of rows in the matrix $\boldsymbol{L}$, denoted by $s$.

@@

To carry out a test of multiple comparisons, it is enough to select the desired contrasts in the lines of the matrix $\boldsymbol{K_1}$ and use these lines as a matrix for specifying the hypotheses of the Wald test. Finally, as usual in tests of multiple comparisons, correction of p-values by means of Bonferroni correction is recommended.

@@

\begin{figure}[h]
\centerline{\includegraphics[scale = 0.9]{FIGURE2}}
\caption{Simulation study results for the regression parameters.\label{fig2}}
\end{figure}

@@

\begin{figure}[h]
\centerline{\includegraphics[scale = 0.9]{FIGURE3}}
\caption{Simulation study results for the dispersion parameters.\label{fig3}}
\end{figure}

@@

$$
g_{r}(\mu_{r}) = \beta_{0r} + \beta_{1r} T1 + \beta_{2r} T2 + \beta_{3r} Probiotic + \\ \beta_{4r} T1*Probiotic + \beta_{5r} T2*Probiotic,
$$

@@

\begin{figure}[h]
\centerline{\includegraphics[scale = 0.9]{FIGURE4}}
\caption{Histogram of Pearson residuals by response.\label{fig4}}
\end{figure}

\begin{figure}[h]
\centerline{\includegraphics[scale = 0.9]{FIGURE5}}
\caption{Pearson residuals versus predicted plot with smooth trend line for each response.\label{fig5}}
\end{figure}

@@

\begin{figure}[h]
\centerline{\includegraphics[width=7in,height=4in]{FIGURE6}}
\caption{Graph of model predicted for each time and group combination.\label{fig6}}
\end{figure}

@@

%\backmatter

\section*{Acknowledgments}

The authors thank the reviewers for their constructive and helpful comments, which greatly improved the article. This study was financed in part by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior – Brasil (CAPES) – Finance Code 001.

@@

\subsection*{Conflict of interest}

The authors declare no potential conflict of interests.

%\section*{Supporting information}

%The following supporting information is available as part of the online article:

%\noindent
%\textbf{Figure S1.}
%{500{\uns}hPa geopotential anomalies for GC2C calculated against the ERA %Interim reanalysis. The period is 1989--2008.}

%\noindent
%\textbf{Figure S2.}
%{The SST anomalies for GC2C calculated against the observations (OIsst).}

%\nocite{*}% Show all bib entries - both cited and uncited; comment this line to view only cited bib entries;

\bibliography{draft-sm-lineu.bib}%

\clearpage

>>>>

% sage_latex_guidelines.tex V1.20, 14 January 2017

\documentclass[Review,sagev,times, doublespace]{sagej}

\usepackage{moreverb,url}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{2016}

@@

\runninghead{Freitas et al}

@@

\author{Lineu Alberto Cavazani de Freitas\affilnum{1}, Ligia de Oliveira Carlos\affilnum{2}, Antônio Carlos Ligocki Campos\affilnum{2} and Wagner Hugo Bonat\affilnum{3}}

\affiliation{
    \affilnum{1}Department of Informatics, Paraná Federal University, Paraná, Brazil\\
    \affilnum{2}Department of Surgical Clinic, Paraná Federal University, Paraná, Brazil\\
    \affilnum{3}Department of Statistics, Paraná Federal University, Paraná, Brazil\\    
}

\corrauth{Lineu Alberto Cavazani de Freitas, Department of Informatics, Paraná Federal University, R. Evaristo F. Ferreira da Costa, 383-391, Jardim das Américas, Curitiba, Paraná, 82590-300, Brazil.}

\email{lineuacf@gmail.com}

\begin{abstract}

@@

\end{abstract}

@@

@@

@@

According to Meinert and Tonascia \cite{meinert1986clinical} ``a clinical trial is a planned experiment designed to assess the efficacy of a treatment in man by comparing the outcomes in a group of patients treated with the test with those observed in a comparable group of patients receiving a control treatment where patients in both groups are enrolled, treated, and followed over the same time period".

@@

Similarly to the aforementioned articles, we are interested in the analysis of multiple outcomes in the context of longitudinal data analysis. The study we shall describe was conducted with the objective of evaluating the effect of the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. This is a clinical trial with two response variables: a score that characterizes compulsion and the number of symptoms presented that characterize addiction. Both responses are limited counts and for analysis purposes have been transformed to unit range using the so-called Min-Max normalization.

@@

It is important to note that the asymptotic normality of estimators of McGLMs was established in Bonat and Jørgensen \cite{Bonat16}, and Wald testing for estimators which are asymptotically normal is a standard approach in asymptotic statistics due to the central limit theorem. So, in this work, our objective is to show how to proceed to perform Wald tests for McGLMs and verify if the expected properties are met.

@@

We present the dataset in \autoref{sec2} 2. In \autoref{sec3} 3 we present the review of the general structure and estimation of the parameters of a McGLM, based on the ideas of Bonat and Jørgensen \cite{Bonat16}. In \autoref{sec4} 4 our proposal is presented with the details of the Wald test to evaluate assumptions about parameters of a McGLM. In \autoref{sec5} 5 we present the results of the performance evaluation of the proposed test based on a simulation study. In the \autoref{sec6} 6 we apply the model and present the main results of the analysis of the data from the clinical study that aims to evaluate the use of probiotics in the control of addiction and binge eating disorder in patients undergoing bariatric surgery. Finally, the main results are discussed in \autoref{sec7} 7, including some directions for future investigations. The data sets and R codes are available online and can be assessed in \href{https://github.com/lineu96/paper-htmcglm}{https://github.com/lineu96/paper-htmcglm}.

@@

\begin{figure}[h]
\centering
\includegraphics[width=29.6pc,height=22.2pc]{FIGURE1.jpeg}
\caption{Exploratory graphic analysis: (a) YFAS histogram, (b) YFAS boxplots as a function of group, (c) YFAS boxplots as a function of moment, (d) BES histogram, (b) BES boxplots as a function of group, (c) BES boxplots as a function of moment. The asterisk in the boxplots indicates the mean.\label{fig1}}
\end{figure}

@@

\noindent where the $g_r()$ are standard link functions that relates the response to the predictor; $\boldsymbol{X}_r$ denotes a $N \times k_r$ design matrix; $\boldsymbol{\beta}_r$ denotes a  $k_r \times 1$ vector of regression parameters. $\boldsymbol{\Sigma}_R \overset{G} \otimes \boldsymbol{\Sigma}_b = \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1, \ldots, \tilde{\boldsymbol{ \Sigma}}_R) (\boldsymbol{\Sigma}_b \otimes \boldsymbol{I}) \mathrm{Bdiag}(\tilde{\boldsymbol{\Sigma}}_1^\top, \ldots, \tilde{\boldsymbol{\Sigma}}_R^\top)$ denotes the generalized Kronecker product \cite{martinez13}, the matrix $\tilde{\boldsymbol{\Sigma}}_r$ denotes the lower triangular matrix of the Cholesky decomposition of the matrix ${\boldsymbol{\Sigma}}_r$. The operator $\mathrm{Bdiag()}$ denotes the block-diagonal matrix and $\boldsymbol{I}$ is an identity matrix $N \times N$.

@@

McGLMs' second-moment assumptions allow us to split the parameter vector into two sets: $\boldsymbol{\theta} = (\boldsymbol{\beta}^{\top}, \boldsymbol{\lambda}^{\top})^ {\top}$. So $\boldsymbol{\beta} = (\boldsymbol{\beta}_1^\top, \ldots, \boldsymbol{\beta}_R^\top)^\top$ is a $K \times 1$ vector of regression parameters and $\boldsymbol{\lambda} = (\rho_1, \ldots, \rho_{R(R-1)/2}, p_1, \ldots, p_R, \boldsymbol{\tau}_1^\top , \ldots, \boldsymbol{\tau}_R^\top)^\top$ is a $Q \times 1$ vector of dispersion parameters. Also, $\mathcal{Y} = (\boldsymbol{Y}_1^\top, \ldots, \boldsymbol{Y}_R^\top)^\top$ denotes the stacked vector of order $NR \times 1$ from the matrix of response variables $\boldsymbol{Y}_{N \times R}$ and $\mathcal{M} = (\boldsymbol{\mu}_1^\top, \ldots, \boldsymbol{\mu}_R^ \top)^\top$ denotes the stacked vector of order $NR \times 1$ of the expected value matrix $\boldsymbol{M}_{N \times R}$.

@@

\noindent In this way, the Wald test statistic to verify the validity of a hypothesis about parameters of a McGLM is given by:

@@

\noindent where $W \sim \chi^2_s$, that is, regardless of the number of parameters in the hypotheses, the test statistic $W$ is a single value that asymptotically follows the $\chi^2$ distribution with degrees of freedom given by the number of constraints, that is, the number of rows in the matrix $\boldsymbol{L}$, denoted by $s$. Some points must be taken into account about the statement $W \sim \chi^2_s$: it's only holds under the null hypothesis; for non-Gaussian data, the statement only holds asymptotically as sample size tends to infinity; and the statement only holds if the matrix $\boldsymbol{L}$ has rank $s$.

@@

To carry out a test of multiple comparisons, it is enough to select the desired contrasts in the lines of the matrix $\boldsymbol{K_1}$ and use these lines as a matrix for specifying the hypotheses of the Wald test. Finally, due to the amount of hypothesis testing in a multiple comparison procedure, correction of p-values by means of Bonferroni correction is recommended.

@@

\begin{figure}[h]
\centering
\includegraphics[width=29.6pc,height=22.2pc]{FIGURE2.jpeg}
\caption{Simulation study results for the regression parameters.\label{fig2}}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=29.6pc,height=22.2pc]{FIGURE3.jpeg}
\caption{Simulation study results for the dispersion parameters.\label{fig3}}
\end{figure}

@@

\begin{equation*}
  \begin{array}{r}
    g_{r}(\mu_{r}) = \beta_{0r} + \beta_{1r} T1 + \beta_{2r} T2 + \beta_{3r} Probiotic + \\ \beta_{4r} T1*Probiotic + \beta_{5r} T2*Probiotic,
  \end{array}
\end{equation*}

@@

\begin{figure}[h]
\centering
\includegraphics[width=29.4pc,height=12.6pc]{FIGURE4.jpeg}
\caption{Histogram of Pearson residuals by response.\label{fig4}}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=29.4pc,height=12.6pc]{FIGURE5.jpeg}
\caption{Pearson residuals versus predicted plot with smooth trend line for each response.\label{fig5}}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=31.5pc,height=18pc]{FIGURE6.jpeg}
\caption{Graph of model predicted for each time and group combination.\label{fig6}}
\end{figure}

@@

\begin{acks}
The authors thank the reviewers for their constructive and helpful comments, which greatly improved the article.
\end{acks}

%\begin{biog}
%To typeset an
%  "Author biography" section.
%\end{biog}

%\begin{biogs}
%To typeset an
%  "Author Biographies" section.
%\end{biogs}

\begin{dci}
The Authors declares that there is no conflict of interest.
\end{dci}

\begin{funding}
This study was
financed in part by the Coordenação de Aperfeiçoamento de Pessoal de Nível Superior – Brasil (CAPES) – Finance Code 001.
\end{funding}

\begin{sm}
R code and data set are available at https://github.com/lineu96/paper-htmcglm
\end{sm}

\bibliographystyle{SageV}
\bibliography{draft-sm-lineu.bib}

@@
